Interactive effects of drought and habitat fragmentation on vital rates of an understory tropical plant

### 2021-01-29 analysis/paper/paper.Rmd
edits, back to Eric

We assessed the effects of drought history on plant vital rates using distributed lag non-linear models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017]. DLNMs model potentially delayed effects of predictor variables by fitting a bi-dimensional dose-lag-response association spline, called a crossbasis function. This crossbasis function allows the effect of a predictor to vary non-linearly through lag time and across predictor intensity. Crossbasis functions were created using the *dlnm* package in R with a maximum lag of 36 months [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020]. This particular maximum lag time was chosen because previous experiments showed that *H. acuminata* individuals recovered from transplant shock after 36 months, so we expected at most 36 months lag time in the response to drought [@brunaExperimentalAssessmentHeliconia2002]. The crossbasis function was fit to the data in the context of a generalized additive model (GAM) using the *mgcv* package in R with restricted maximum likelihood [@woodGeneralizedAdditiveModels2017]. In the process of fitting the model, the penalty on the crossbasis smooth (and other smoothed terms) is optimized such that more linear shapes are favored unless the data supports non-linearity.

@@

Where $f_1(x_i)$ is a smooth function of plant size [ ln(height $\times$ shoot number) ], fit using a penalized cubic regression spline, $f_2()$ is the crossbasis function in which $d_{i,t}$ is the SPEI value during the census month of an observation (February) and $d_{i,t-l}$ is the SPEI $l$ months prior (see Gasparrini et al. 2017 for details). Penalized cubic regression splines were used for both dimensions of the crossbasis function, with 35 knots for the lag dimension and 3 knots for the drought response dimension to restrict the fitted response to drought to bimodal at most complex. A random effect of plot ID on the intercept is represented by $a_j$. For plant growth the response was plant size (height $\times$ shoot number) in year t+1 modeled with a scaled t family error distribution and a log link function ($g(\cdot) = log(\cdot)$). A scaled t family was chosen because residuals were leptokurtic with a gaussian error structure. The data on number of infloresences was converted to binary (1 for ≥1 inflorescence, 0 for no inflorescence). For survival and flowering as a response, we used a binomial family error distribution with a logit link.

>>>>

We assessed the effects of drought history on plant vital rates using distributed lag non-linear models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017]. DLNMs model potentially delayed effects of predictor variables by fitting a bi-dimensional dose-lag-response association spline, referred to as the crossbasis function. This function allows the effect of a predictor to vary non-linearly through lag time and across predictor intensity. Crossbasis functions were created using the *dlnm* package with a maximum lag of 36 months [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020]. This particular maximum lag time in respont to drought was chosen because *H. acuminata* individuals transplanted to forest fragments typically recovered from transplant shock within 36 months [@brunaExperimentalAssessmentHeliconia2002]. The crossbasis function was fit to the data in the context of a generalized additive model (GAM) using the *mgcv* package with restricted maximum likelihood [@woodGeneralizedAdditiveModels2017]. In the process of fitting the model, the penalty on the crossbasis smooth (and other smoothed terms) is optimized such that more linear shapes are favored unless the data supports non-linearity.

@@

Where $f_1(x_i)$ is a smooth function of plant size [ ln(height $\times$ shoot number) ], fit using a penalized cubic regression spline, $f_2()$ is the crossbasis function in which $d_{i,t}$ is the SPEI value during the census month of an observation (February) and $d_{i,t-l}$ is the SPEI $l$ months prior (see Gasparrini et al. 2017 for details). Penalized cubic regression splines were used for both dimensions of the crossbasis function, with 35 knots for the lag dimension and 3 knots for the drought response dimension to restrict the fitted response to drought to bimodal at most complex. A random effect of plot ID on the intercept is represented by $a_j$. For plant growth the response was plant size (height $\times$ shoot number) in year t+1 modeled with a scaled t family error distribution and a log link function ($g(\cdot) = log(\cdot)$). A scaled t family was chosen because residuals were leptokurtic with a Gaussian error structure. The data on number of infloresences was converted to binary (1 for ≥1 inflorescence, 0 for no inflorescences). When survival and flowering were the response variables we used a binomial family error distribution with a logit link function.
